---
title: "Praxis-BGM R Wrapper"
author: "Qiran Jia"
date: "2026-04-16"
output:
  html_document: default
---

## Overview

This document shows how to call the updated Python `praxis_bgm` package from R
with `reticulate`. It is written for the current installable API:

- Python package: `praxis_bgm`
- Main class: `praxis_bgm.Praxis_BGM`
- Assumed environment: `Praxis_env`

Compared with the older wrapper, the updated Python API now uses:

- `prior_weights` instead of `prior_pis`
- `init_pis` for initialization weights
- `freeze_A_zeros` instead of the older mask-enforcement arguments
- additional runtime controls such as `data_precision_int`,
  `likelihood_temp`, `rho_prec`, `rho_mu`, and `elbo_eval_freq`

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)

library(reticulate)

use_condaenv("Praxis_env", required = TRUE)
```

```{r py-config}
py_config()
```

## R Wrapper

```{r wrapper}
praxis_bgm_fit <- function(
  data,
  K,
  seed = 0L,
  prior_mus = NULL,
  prior_Sigmas = NULL,
  prior_weights = NULL,
  init_mus = NULL,
  init_covs = NULL,
  init_pis = NULL,
  beta = 1e-3,
  tol = 1e-4,
  max_iters = 1000L,
  verbose = TRUE,
  sparse_A = NULL,
  cluster_A = NULL,
  freeze_A_zeros = FALSE,
  prior_mus_variance = 1.0,
  num_samples = 100L,
  data_precision_int = NULL,
  likelihood_temp = 1.0,
  rho_prec = 0.05,
  rho_mu = 1.0,
  elbo_eval_freq = 10L,
  num_iters = 50L,
  batch_size = NULL,
  early_stop = FALSE,
  patience = 2L
) {
  if (!reticulate::py_module_available("praxis_bgm")) {
    stop("The Python module `praxis_bgm` is not available in `Praxis_env`.")
  }
  if (!reticulate::py_module_available("jax")) {
    stop("The Python module `jax` is not available in `Praxis_env`.")
  }
  if (missing(data)) {
    stop("`data` is required and must be a numeric matrix or data frame.")
  }
  if (!is.numeric(K) || length(K) != 1L || is.na(K) || K < 2 || K != as.integer(K)) {
    stop("`K` must be a single integer greater than or equal to 2.")
  }

  data_matrix <- if (is.data.frame(data)) {
    data.matrix(data)
  } else {
    as.matrix(data)
  }

  if (!is.matrix(data_matrix) || length(dim(data_matrix)) != 2L) {
    stop("`data` must be a two-dimensional matrix or data frame.")
  }
  if (!is.numeric(data_matrix)) {
    stop("`data` must contain numeric values.")
  }
  if (any(!is.finite(data_matrix))) {
    stop("`data` must contain only finite values.")
  }

  n_obs <- nrow(data_matrix)
  if (is.null(batch_size)) {
    batch_size <- min(50L, n_obs)
  }
  if (!is.numeric(batch_size) || length(batch_size) != 1L || batch_size < 1) {
    stop("`batch_size` must be a positive integer.")
  }
  if (batch_size > n_obs) {
    stop("`batch_size` cannot exceed the number of rows in `data`.")
  }

  np <- reticulate::import("numpy", delay_load = TRUE, convert = FALSE)
  jax <- reticulate::import("jax", delay_load = TRUE, convert = FALSE)
  praxis <- reticulate::import("praxis_bgm", delay_load = TRUE, convert = FALSE)

  as_numpy <- function(x, dtype = "float32") {
    if (is.null(x)) {
      return(NULL)
    }
    if (inherits(x, "python.builtin.object")) {
      return(x)
    }
    np$array(x, dtype = dtype)
  }

  py_array_to_r <- function(x) {
    if (is.null(x)) {
      return(NULL)
    }
    reticulate::py_to_r(np$asarray(x))
  }

  rng_key <- jax$random$PRNGKey(as.integer(seed))

  model <- praxis$Praxis_BGM(
    rng_key = rng_key,
    K = as.integer(K),
    prior_mus = as_numpy(prior_mus),
    prior_Sigmas = as_numpy(prior_Sigmas),
    prior_weights = as_numpy(prior_weights),
    init_mus = as_numpy(init_mus),
    init_covs = as_numpy(init_covs),
    init_pis = as_numpy(init_pis),
    beta = as.numeric(beta),
    tol = as.numeric(tol),
    max_iters = as.integer(max_iters),
    verbose = isTRUE(verbose),
    sparse_A = as_numpy(sparse_A),
    cluster_A = as_numpy(cluster_A),
    freeze_A_zeros = isTRUE(freeze_A_zeros),
    prior_mus_variance = as.numeric(prior_mus_variance),
    num_samples = as.integer(num_samples),
    data_precision_int = if (is.null(data_precision_int)) NULL else as.integer(data_precision_int),
    likelihood_temp = as.numeric(likelihood_temp),
    rho_prec = as.numeric(rho_prec),
    rho_mu = as.numeric(rho_mu),
    elbo_eval_freq = as.integer(elbo_eval_freq)
  )

  model$fit(
    as_numpy(data_matrix),
    num_iters = as.integer(num_iters),
    batch_size = as.integer(batch_size),
    early_stop = isTRUE(early_stop),
    patience = as.integer(patience)
  )

  posteriors <- model$get_posteriors(as_numpy(data_matrix))
  prediction <- model$predict(as_numpy(data_matrix))

  posterior_mus <- py_array_to_r(posteriors[[0]])
  posterior_covs <- py_array_to_r(posteriors[[1]])
  posterior_pis <- py_array_to_r(posteriors[[2]])
  responsibilities <- py_array_to_r(posteriors[[3]])

  assignments_zero_based <- reticulate::py_to_r(prediction[[0]])
  learned_weights <- reticulate::py_to_r(prediction[[1]])
  assignments <- as.integer(assignments_zero_based) + 1L

  list(
    assignments = assignments,
    assignments_zero_based = assignments_zero_based,
    learned_weights = learned_weights,
    posterior_mus = posterior_mus,
    posterior_covs = posterior_covs,
    posterior_pis = posterior_pis,
    responsibilities = responsibilities,
    model_summary = reticulate::py_to_r(model$get_model_summary()),
    elbo_history = reticulate::py_to_r(model$elbo_history),
    model = model
  )
}
```

## Self-contained Example

```{r example-data}
set.seed(123)

n_per_cluster <- 40

cluster_1 <- cbind(
  rnorm(n_per_cluster, mean = -2.0, sd = 0.45),
  rnorm(n_per_cluster, mean = -1.8, sd = 0.50)
)

cluster_2 <- cbind(
  rnorm(n_per_cluster, mean = 2.1, sd = 0.50),
  rnorm(n_per_cluster, mean = 1.9, sd = 0.45)
)

X <- rbind(cluster_1, cluster_2)
colnames(X) <- c("feature_1", "feature_2")

prior_mus <- rbind(
  c(-1.9, -1.9),
  c(2.0, 2.0)
)

prior_Sigmas <- array(0, dim = c(2, 2, 2))
prior_Sigmas[1, , ] <- diag(c(1.0, 1.0))
prior_Sigmas[2, , ] <- diag(c(1.0, 1.0))

true_labels <- c(rep(1L, n_per_cluster), rep(2L, n_per_cluster))
```

```{r example-fit}
fit <- praxis_bgm_fit(
  data = X,
  K = 2L,
  seed = 42L,
  prior_mus = prior_mus,
  prior_Sigmas = prior_Sigmas,
  prior_weights = c(0.5, 0.5),
  beta = 1e-3,
  tol = 1e-4,
  max_iters = 60L,
  num_iters = 8L,
  batch_size = 40L,
  num_samples = 8L,
  data_precision_int = 1L,
  likelihood_temp = 1.0,
  rho_prec = 0.08,
  rho_mu = 1.0,
  elbo_eval_freq = 1L,
  early_stop = TRUE,
  patience = 2L,
  verbose = FALSE
)
```

## Results

```{r results}
fit$assignments[1:10]
fit$learned_weights
fit$model_summary
tail(fit$elbo_history, 3)
table(True = true_labels, Predicted = fit$assignments)
```

## Notes

```{r notes, echo=FALSE}
cat(
  paste(
    "- `assignments` are returned as 1-based cluster labels for R users.",
    "- `assignments_zero_based` preserves the raw Python labels from `model.predict()`.",
    "- `posterior_mus`, `posterior_covs`, `posterior_pis`, and `responsibilities` are converted back into R objects.",
    "- `model` remains a live Python object, so you can still call methods such as `fit$model$BF_selection(...)` through `reticulate`.",
    sep = "\n"
  )
)
```
